---
title: "isoSCAN Package Vignette"
author: "Jordi Capellades"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{isoSCAN Package Vignette}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---
```{r, echo = FALSE}
muffleError <- function(x,options) {}
knitr::knit_hooks$set(error=muffleError)
knitr::opts_chunk$set(collapse = TRUE, purl = FALSE)
```
## Introduction
This vignette contains a basic walkthrough of the functionalities of the `isoSCAN` package.

This package requires a specific input format for the _compound table_ to be used in _autoQ_ _formulaTable_ argument, in which we __must__ find different columns in no specific order:

* __CompoundName__ is the name of the compound or metabolite quantified

* __mz__ of the monoisotopic ion. The other isotopologue ions are calculated by _autoQ_ function

* __RT__ retention time value in seconds

* __Formula__ of the compound. This formula must match the derivatized _Formula_ including derivatization modifications.

* __NumAtoms__ determining the number of compounds to be quantified

* Other columns will be ignored

The rest of the functions are used for processing the raw data for either quantification or plotting. The package currently contains the following functions.

* Quantifcation:

    * `autoQ`

##### TODO: Complete function list
<!-- * Value transformation -->

<!--     * `QTransform` -->

<!-- * Plotting  -->

<!--     * `rawPlot` -->

<!--     * `meanRawPlot` -->

<!--     * `metBarPlot` -->

## Creating the formulaTable
Before starting with the file processing, we need to load the _compound table_ as the _formulaTable_ data frame. This can be done either with `read.table` or `read.csv` functions. Make sure that the file contains it contains the columns as listed above.

The package includes examples for both Low- and High-Resolution.

```{r,message = FALSE,warning = FALSE}
library(isoSCAN)
data("formulaTables")

# Low-Resolution (e.g. nominal mass accuracy)
formulaTable_lowres <- formulaTables[which(formulaTables$Instrument=="Quadrupole"),]

formulaTable_lowres
```


```{r}
# High-Resolution (Orbitrap, or qTOF)
formulaTable_orbi <- formulaTables[which(formulaTables$Instrument=="Orbitrap"),]

formulaTable_orbi
```


## Processing mz(X)ML files
First, you will have to transform the raw data from vendor format into mz(X)ML format using Proteowizard MSconvert (or similar tools), so they can be read by the `mzR` R package.
* In the case of Low-resolution. Transform the data mantaining profile format.
* In the case of High-resolution, please use centroiding (e.g. _peakPicking true_ in msconvert)
Then, we need to locate the folder in which these files are found and list them in a vector.

```{r,eval=FALSE}
#run this: setwd("./mydatafolder")
SampleFiles  <- list.files(pattern=".mz(X)?ML")
```

This package also includes example mzML data files to be used as an example:
```{r,eval=FALSE}
# Low-resolution files
SampleFiles_lowres <- list.files(system.file("extdata",package = "isoSCAN"),full.names = T,pattern = "lowres")

#High-resolution files
SampleFiles_orbi <- list.files(system.file("extdata",package = "isoSCAN"),full.names = T,pattern = "orbi")
```

Now we can call `autoQ` function that will process the files and look for the isotopologues for each compound found in the `formulaTable`.
Additionally, other parameters need to be indicated as stated in _help(autoQ)_. This parameters refer to peak width and number of scans recorded, together with signal-to-noise ratio and mass error.

In the case of low-resolution data. Please remember to use them in _Profile_ format as it eases the process of peak finding. 

```{r,eval=FALSE}
# Low-Resolution
integrations <- autoQ(SampleFiles=SampleFiles_lowres,
                      formulaTable=formulaTable_lowres,
                      resolution = 1,  # low resolution
                      minscans = 6,
                      SNR = 3,
                      mzerror = 0.1,
                      RTwin = 5,
                      maxwidth = 4,
                      minwidth = 1,
                      massdiff = 1.003355)


head(integrations)
```


###### TODO: explain enviPat

```{r,eval=FALSE}
data(isotopes, package="enviPat")
isotopes[isotopes$isotope=="13C",] # both rows required

# High-Resolution
integrations <- autoQ(SampleFiles=SampleFiles_orbi,
                      formulaTable=formulaTable_orbi,
                      resolution = c(1e5,200),  # orbi resolution parameters
                      minscans = 6,
                      SNR = 3,
                      maxppm=5,
                      RTwin = 5,
                      maxwidth = 4,
                      minwidth = 1,
                      isotopes=isotopes, 
                      labelatom="13C")
```

This processes each file independently, looking for "good-shape" peaks and obtaining both the area and max intensity scan (Maxo) for each isotopologue, if the area cannot be calculated (due to noise or peak shape) then only the Maxo is returned.

Once finished, we can plot them or transform the values for exportation. 

###### TODO: Adapt plotting to new version 0.1 >> 0.9

## Plotting results and value transformation
The `metBarPlot` function is designed to plot values in a barplot including standard deviation error bars. The arguments for value are the following:

* _groups_ Sample groups. This should be a vector with the groups of experiments, matching the same order in `list.files(pattern=".mzXML")`. This vector can be created using `gsub` function and others. See the example:

* _val.to.plot_ Values to use for plotting. Either "Area" or "Maxo".

```{r,eval=FALSE}
mygroups <- isoSCAN:::rmfileExt(SampleFiles_orbi,"\\.mzML")
mygroups <- gsub(".*_","",mygroups)
```


```{r,eval=FALSE}
metBarPlot(x=integrations, groups = mygroups, val.to.plot="Area")
#if you want to save the plots to a file:
metBarPlot(x=integrations, groups = mygroups, val.to.plot="Area",
					 topdf="./metBarPlot_results.pdf",height=10,width=18,pointsize=16)
#modify height, width and pointsize accordingly to fit your output size

```

* _ylabel_ The text that should be shown in the Y axis (i.e. Intensity, Area). Otherwise, _val.to.plot_ value will be used. 

`metBarPlot` can also digest the data frame produced by `QTransform`. Read the function help for further information. 

## Extra: Raw data plotting
`rawPlot` and `meanRawPlot` functions should be used for quality control purposes. They are useful to check for moving peaks, noisy spots or saturated peaks.
The first function will print a "spectra heatmap", m/z and retention time in the x,y-axis respectively, with points coloured depending on the scan intensity value. 

```{r, eval=FALSE}
# Example plot from first file and first compound
rawPlot(SampleFiles=lf[1],formulaTable=formulaTable[1,],RTwin=5)
```

While the `rawPlot` function will run through each file and plot the raw spectra for all compounds in a single pdf file, `meanRawPlot` will calculate the average spectra for all files and generate a single plot for each compound.

```{r, eval=FALSE}
# Example plot for the first compound, note that now we use ALL files
meanRawPlot(SampleFiles=lf,formulaTable=formulaTable[1,],RTwin=5)
```

Both functions contain the `topdf` argument, that is encouraged to be used in order to save the files into _PDF format_ instead of being shown in the R plotting default device. In the case of `meanRawPlot` indicate the name of the output PDF file desired (`topdf=/plot_folder/mean_raw_spectra.pdf`), whereas in the case of `rawPlot` should be used as: `topdf="C:/Users/User/Desktop/plot_folder` and all plots generated will be saved into *plot_folder* mantaining the sample name for the pdf file name.
